---
title: "KWF_Metals_2019_2020"
output:
  html_document: 
    df_print: paged
    fig_width: 10
    fig_height: 6
    fig_caption: yes
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: false
editor_options: 
  chunk_output_type: inline
---

Document last updated `r Sys.time()` by Benjamin Meyer (ben@kenaiwatershed.org)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# clear environment
rm(list=ls())

# load packages
library(tidyverse)
library(lubridate)
library(readr)
library(readxl)
library(writexl)
library(hms)
library(plotly)
library(DT)
library(xlsx)
library(leaflet)
library(DT)
library(ggpubr)
```

<br>

This draft document contains preliminary data explorations of 2019-2020 copper and zinc water quality data from the Kenai River Watershed.

<br>

Notes:

* Read in data directly from the two AWQMS-formatted documents for 2019 and 2020 found on the Kenai Watershed Forum local server.

<br>

***

Read in and prepare data
```{r}
# read in excel files
dat19 <- read_excel("data/2019_KWF_AWQMS_Results.xlsx", sheet = "AWQMS Data Template") %>%
  mutate_all(as.character)
dat20 <- read_excel("data/2020_KWF_AWQMS_Results.xlsx", sheet = "AWQMS Data Template", 
                    skip = 1) %>%
  mutate_all(as.character)

# specify col names to retain
vars <- c("Monitoring Location ID",
          "Activity Media Name",
          "Activity Start Date",
          "Activity Latitude",
          "Activity Longitude",
          "Activity Type",
          "Characteristic Name",
          "Result Value",
          "Result Unit",
          "Result Qualifier",
          "Result Value Type",
          "Result Status ID")

# create new variable names
nm <- c("Location",
        "Medium",
        "Date",
        "Latitude",
        "Longitude",
        "Activity",
        "Measurememnt", 
        "Value",
        "Unit",
        "Qualifier",
        "Value_Type",
        "Status")

# join the two years
dat <- bind_rows(dat19,dat20) %>%
  select(all_of(vars)) %>%
  filter(!is.na(`Monitoring Location ID`))

# rename columns
colnames(dat) <- nm

# remove unneeded items
rm(dat19,dat20,nm,vars)

```

<br>


***

### Locations
```{r}
(sites <- dat %>%
  select(Location,Latitude,Longitude) %>%
  distinct() %>%
  mutate(Latitude = as.numeric(Latitude),
         Longitude = as.numeric(Longitude)))
```

<br>

Show locations on map
```{r}
leaflet(data = sites) %>% 
  addTiles() %>%
  addMarkers(~Longitude, ~Latitude, popup = ~as.character(Location))
```

<br>

### Comments

Note: The locations shown above are inclusive of all known data found on the two AWQMS-formatted documents on the Kenai Watershed Forum server.  The locations appear to include solely the supplemental Copper and Zinc-specific sampling event sites of 2019-2020, and do not include the other standard biannual KRBWQM sites.

To do: 

* Ensure that all 2019 - 2020 has been submitted to DEC, including the KRBWQM data
* Locate, format, and submit any missing data to DEC if not already present in their archive
* For generating the 2019 - 2020 report, we will use the tables prepared in the Excel document, "AllData_ZnCuCaMg_19-20.xlsx"
* See main RPubs document for a continuation of this analysis: https://rpubs.com/bmeyer/cu_zn_2019_2020 , using data from "AllData_ZnCuCaMg_19-20.xlsx"


