---
title: "KWF_metals_2020"
output:
  html_document: 
    df_print: paged
    fig_width: 10
    fig_height: 6
    fig_caption: yes
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: false
editor_options: 
  chunk_output_type: inline
---

Document last updated `r Sys.time()` by Benjamin Meyer (ben@kenaiwatershed.org)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# clear environment
rm(list=ls())

# load packages
library(tidyverse)
library(lubridate)
library(readr)
library(readxl)
library(writexl)
library(hms)
library(plotly)
library(DT)
library(xlsx)
library(leaflet)
library(DT)
library(ggpubr)
```

<br>

This draft document contains preliminary data explorations of 2019-2020 copper and zinc water quality data from the Kenai River Watershed.

<br>

Notes:

* In the future it will be much better to read in final data directly from the EPA institutional archive, in its most static version.  I found Cu, Zn, Mg, and Ca data pre-2013 at https://www.waterqualitydata.us/portal/, but did not yet find 2013-2020 data there.

* Specific query used: https://www.waterqualitydata.us/portal/#countrycode=US&statecode=US%3A02&countycode=US%3A02%3A122&sampleMedia=water&sampleMedia=Water&characteristicType=Inorganics%2C%20Major%2C%20Metals&characteristicType=Inorganics%2C%20Minor%2C%20Metals&mimeType=csv&dataProfile=narrowResult

* Where are the rest of the years archived? Do they need to be submitted AQMS?  Maybe use a different query?

* In cases 2019 where hardness data not available, data from nearest sampling event was used instead.

<br>

***

### Data Read in and preparation

Read in results columns directly from tabs "FINALREPORT_ZnCuSampling_19-20" and "FINALREPORT_Baseline_19-20" in the Excel document "AllData_ZnCuCaMg_19-20.xlsx" prepared by MH in Fall 2020
```{r}
# specify file paths and sheet names
workbook <- "data/AllData_ZnCuCaMg_19-20.xlsx"
zn_cu_fnl <- "FINALREPORT_ZnCuSampling_19-20"
baseline_fnl <- "FINALREPORT_Baseline_19-20"

# read in tabs
zn_cu_fnl <- read_excel(workbook, sheet = zn_cu_fnl, skip = 4) 
baseline_fnl <- read_excel(workbook, sheet = baseline_fnl, skip = 4)

# join dataframes
tbl <- bind_rows(zn_cu_fnl,baseline_fnl)

# assign new column names
vars <- c("site","rm","date","waterbody_type","hardness_mgL","copper_ugL","copper_ccc_ugL","zinc_ugL","zinc_ccc_ugL")
colnames(tbl) <- vars

# create numerical columns and clean data table
tbl <- tbl%>%
  select(-rm) %>%
  filter(!is.na(date)) %>%
  fill(site) %>%
  
  # hardness
  separate(hardness_mgL, sep = " ", into = c("hardness_mgL_1","hardness_mgL_2"), remove = F) %>%
  mutate(hardness_mgL_2 = gsub("[()]", "", hardness_mgL_2)) %>%
  
  # copper
  mutate(copper_ugL_x = gsub("[()]", "", copper_ugL)) %>%
  mutate(copper_ugL_x = gsub("J", "", copper_ugL_x)) %>%
  mutate(copper_ugL_x = gsub("U", "", copper_ugL_x)) %>%
  mutate(copper_ugL_x = str_replace(copper_ugL_x, "\\s", "|")) %>% 
  ### replace first white space with consistent character and split on that character
  separate(copper_ugL_x, into = c("copper_ugL_1", "copper_ugL_2"), sep = "\\|") %>%
  
  # copper CCC
  mutate(copper_ccc_ugL_x = gsub("[()]", "", copper_ccc_ugL)) %>%
  mutate(copper_ccc_ugL_x = gsub("J", "", copper_ccc_ugL_x)) %>%
  mutate(copper_ccc_ugL_x = gsub("U", "", copper_ccc_ugL_x)) %>%
  mutate(copper_ccc_ugL_x = str_replace(copper_ccc_ugL_x, "\\s", "|")) %>% 
  ### replace first white space with consistent character and split on that character
  separate(copper_ccc_ugL_x, into = c("copper_ccc_ugL_1", "copper_ccc_ugL_2"), sep = "\\|") %>%
  
  # zinc
  mutate(zinc_ugL_x = gsub("[()]", "", zinc_ugL)) %>%
  mutate(zinc_ugL_x = gsub("J", "", zinc_ugL_x)) %>%
  mutate(zinc_ugL_x = gsub("U", "", zinc_ugL_x)) %>%
  mutate(zinc_ugL_x = str_replace(zinc_ugL_x, "\\s", "|")) %>% 
  ### replace first white space with consistent character and split on that character
  separate(zinc_ugL_x, into = c("zinc_ugL_1", "zinc_ugL_2"), sep = "\\|") %>%
  
  # zinc CCC
  mutate(zinc_ccc_ugL_x = gsub("[()]", "", zinc_ccc_ugL)) %>%
  mutate(zinc_ccc_ugL_x = gsub("J", "", zinc_ccc_ugL_x)) %>%
  mutate(zinc_ccc_ugL_x = gsub("U", "", zinc_ccc_ugL_x)) %>%
  mutate(zinc_ccc_ugL_x = str_replace(zinc_ccc_ugL_x, "\\s", "|")) %>% 
  ### replace first white space with consistent character and split on that character
  separate(zinc_ccc_ugL_x, into = c("zinc_ccc_ugL_1", "zinc_ccc_ugL_2"), sep = "\\|") %>%
  
  # select desired columns
  select("site","date","waterbody_type",
         "hardness_mgL_1","hardness_mgL_2",
         "copper_ugL_1","copper_ugL_2","copper_ccc_ugL_1","copper_ccc_ugL_2",
         "zinc_ugL_1","zinc_ugL_2","zinc_ccc_ugL_1","zinc_ccc_ugL_2") %>%
  
  # transform numbers to numeric column classes
  mutate_at(c("hardness_mgL_1","hardness_mgL_2",
         "copper_ugL_1","copper_ugL_2","copper_ccc_ugL_1","copper_ccc_ugL_2",
         "zinc_ugL_1","zinc_ugL_2","zinc_ccc_ugL_1","zinc_ccc_ugL_2"), 
         as.numeric) %>%
  
  # modify site names
  mutate(site = as.factor(str_replace_all(site," ","_"))) %>%
  mutate(site = as.factor(str_replace_all(site,"-","_"))) %>%
  mutate(site = as.factor(str_replace_all(site,"'","")))

```

<br>

Calculate RDL values for Zn and Cu duplicate samples
```{r}
tbl <- tbl %>%
  ## copper rdl
  mutate(cu_rdl = paste(sprintf((abs(copper_ugL_1 - copper_ugL_2) / ((copper_ugL_1 + copper_ugL_2) / 2)*100), fmt = '%#.2f'),"%")) %>%
  mutate(cu_rdl = str_replace(cu_rdl, "NA %", "")) %>%
  
  # zinc rdl
  mutate(zn_rdl = paste(sprintf((abs(zinc_ugL_1 - zinc_ugL_2) / ((zinc_ugL_1 + zinc_ugL_2) / 2)*100), fmt = '%#.2f'),"%")) %>%
  mutate(zn_rdl = str_replace(zn_rdl, "NA %", ""))
  
```

<br>

***

### Table

```{r}
# read in coordinates
coords <- read_excel("data/Sampling sites 2020.xlsx", sheet = "BM_format_Sites2020") %>%
  rename(site = Site) %>%
  transform(site = as.factor(site))

# join coordinates to data
tbl <- left_join(tbl,coords)

# present table
datatable(tbl,
          filter = 'top', 
          options = list(pageLength = 5, autoWidth = TRUE))

```

<br>

***

### Map

Show sampling sites and raw data on an ArcGIS Online map (also at https://arcg.is/0rKbbr)
```{r}
# used arcgis online from csv exported here
write.csv(tbl, "output/CuZn_2019_2020_results.csv", row.names = F)

# viewable at https://arcg.is/0rKbbr

```

<br>

2019 - 2020 Cu and Zn sampling sites

<br>

<style>.embed-container {position: relative; padding-bottom: 80%; height: 0; max-width: 100%;} .embed-container iframe, .embed-container object, .embed-container iframe{position: absolute; top: 0; left: 0; width: 100%; height: 100%;} small{position: absolute; z-index: 40; bottom: 0; margin-bottom: -15px;}</style><div class="embed-container"><iframe width="500" height="400" frameborder="0" scrolling="no" marginheight="0" marginwidth="0" title="Copper and Zinc Sampling 2019-2020" src="//www.arcgis.com/apps/Embed/index.html?webmap=b412c05eb1b0492f95f4667e7c86c7c5&extent=-151.3062,60.4157,-150.5832,60.6518&home=true&zoom=true&previewImage=false&scale=true&search=true&searchextent=true&legend=true&basemap_gallery=true&disable_scroll=true&theme=light"></iframe></div>

<br>

***

### Plots

Prep tables for plotting
```{r}
# prep table

# strategy to prep for ggplot (e.g. "long form": break up in to minimum tables and use joins to rebuild min rows?
tbl <- tbl %>%
  select(site, date, waterbody_type, 
         starts_with(c("copper","cu","zinc","zn")),
         Lat, Long, BM_Notes) 

# rdl table
rdl_tbl <- tbl %>% 
  select(site,date,contains("rdl")) %>%
  filter(zn_rdl != "") %>%
  pivot_longer(contains("rdl"),names_to = "rdl", values_to = "rdl_val") %>%
  mutate(metal = gsub("_rdl","",rdl)) %>%
  select(-rdl)

# ccc table
ccc_tbl <- tbl %>%
  select(site,date,contains("ccc")) %>%
  pivot_longer(contains("ccc"), values_to = "ccc_val") %>% 
  filter(!is.na(ccc_val)) %>%
  mutate(metal = gsub("copper","cu",name)) %>%
  mutate(metal = gsub("zinc","zn",metal)) %>%
  mutate(metal = gsub("_ccc_ugL_1","",metal)) %>%
  mutate(metal = gsub("_ccc_ugL_2","",metal)) %>%
  select(-name)

# metal values observations
met_obs <- c("copper_ugL_1","copper_ugL_2","zinc_ugL_1","zinc_ugL_2")
met_tbl <- tbl %>%
  select(site,date,all_of(met_obs)) %>%
  pivot_longer(met_obs, names_to = "metal", values_to = "met_val") %>%
  filter(!is.na(met_val)) %>%
  mutate(metal = gsub("copper","cu",metal)) %>%
  mutate(metal = gsub("zinc","zn",metal)) %>%
  mutate(metal = gsub("_ugL_1","",metal)) %>%
  mutate(metal = gsub("_ugL_2","",metal)) %>%
  distinct()

  
# join tables in long form
tbl2 <- inner_join(met_tbl,ccc_tbl) %>%
  left_join(rdl_tbl) %>%
  left_join(coords) %>%
  arrange(metal,met_val)

```

<br>

Summarise the range of meyal and CCC values observed
```{r}
tbl2 %>%
  group_by(site,metal) %>%
  summarise(met_max = max(met_val),
            met_min = min(met_val),
            ccc_max = max(ccc_val),
            ccc_min = min(ccc_val)) %>%
  arrange(metal, met_max)
```



<br>


#### Tributary site plots
```{r}
# specify facet orders
## tributaries
trib_order <- c("No_Name_Creek",
                "Beaver_Creek",
                "Slikok_Creek",
                "Soldotna_Creek",
                "Funny_River")
tbl2_trib <- tbl2 %>% filter(Habitat == "Tributary")
tbl2_trib$site <- factor(tbl2_trib$site, levels = trib_order)


# specify common ggplot theme here


# create separate charts with different y axes for each metal
cu_trib <- tbl2_trib%>%
  filter(metal == "cu") %>%
  ggplot() +
  geom_point(aes(as.factor(date),as.numeric(met_val), color = metal)) +
  scale_color_manual(values = c("cu" = "red")) +
  geom_point(aes(as.factor(date),as.numeric(ccc_val)), shape = 95, size = 3) +
  facet_grid(.~site, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        strip.background.x = element_blank(),
        strip.text.x = element_text(face = "bold"),
        legend.position = "none") +
  xlab("") +
  ylab("") +
  ylim(0,10)
cu_trib <-  annotate_figure(cu_trib,right = text_grob("Cu", color = "Red", 
                                 face = "bold", size = 14, rot = 0))

zn_trib <- tbl2_trib %>%
  filter(metal == "zn") %>%
  ggplot() +
  geom_point(aes(as.factor(date),as.numeric(met_val))) +
  geom_point(aes(as.factor(date),as.numeric(ccc_val)), shape = 95, size = 3) +
  facet_grid(.~site, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
        strip.background = element_blank(),
        strip.text.x = element_blank()) +
  xlab("") +
  ylab("") +
  ylim(0,175)
zn_trib <-  annotate_figure(zn_trib,right = text_grob("Zn", color = "black", 
                                 face = "bold", size = 16, rot = 0))

# arrange 2 charts together
p <- ggarrange(cu_trib,zn_trib,nrow = 2)
annotate_figure(p,
                top = text_grob("Kenai River Tributary Sites", 
                                color = "black", face = "bold", size = 18),
                left = text_grob("Concentration (ug/L)", color = "black", 
                                 face = "bold", size = 16, rot = 90),
                fig.lab = "— = CCC",
                fig.lab.pos = "bottom.right",
                fig.lab.face = "bold")

# note: used ggarrange rather than ggplot facet_grid for overall because y-axis scales were different

```

<br>

#### Mainstem site plots
```{r}
## mainstem
ms_order <- c("City_of_Kenai_Docks",
              "Cunningham_Park",
              "Upstream_of_Beaver_Creek",
              "Pillars",
              "Poachers_Cove",
              "Slikok_Creek_Kenai_River_Confluence",
              "Soldotna_Bridge",
              "Swiftwater_Park",
              "Jims_Landing",
              "Skilak_Lake_Outlet")

# shorten site names
ms_sites <- c("City_of_Kenai_Docks" = "CoK_Docks",
              "Cunningham_Park" = "C_Park",
              "Upstream_of_Beaver_Creek" = "Upstream_of_Beaver",
              "Pillars" = "Pillars",
              "Poachers_Cove" = "Poachers",
              "Slikok_Creek_Kenai_River_Confluence" = "Slikok",
              "Soldotna_Bridge" = "S_Bridge",
              "Swiftwater_Park" = "Swift",
              "Jims_Landing" = "Jims",
              "Skilak_Lake_Outlet" = "Skilak")

tbl2_ms <- tbl2 %>% filter(Habitat == "Mainstem")
tbl2_ms$site <- factor(tbl2_ms$site, levels = ms_order)


# create separate charts with different y axes for each metal
cu_ms <- tbl2_ms%>%
  filter(metal == "cu") %>%
  ggplot() +
  geom_point(aes(as.factor(date),as.numeric(met_val), color = metal)) +
  scale_color_manual(values = c("cu" = "red")) +
  geom_point(aes(as.factor(date),as.numeric(ccc_val)), shape = 95, size = 3) +
  facet_grid(.~site, scales = "free_y", labeller = as_labeller(ms_sites)) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        strip.background.x = element_blank(),
        strip.text.x = element_text(face = "bold"),
        legend.position = "none") +
  xlab("") +
  ylab("") +
  ylim(0,10)
cu_ms <-  annotate_figure(cu_ms,right = text_grob("Cu", color = "Red", 
                                 face = "bold", size = 14, rot = 0))

zn_ms <- tbl2_ms %>%
  filter(metal == "zn") %>%
  ggplot() +
  geom_point(aes(as.factor(date),as.numeric(met_val))) +
  geom_point(aes(as.factor(date),as.numeric(ccc_val)), shape = 95, size = 3) +
  facet_grid(.~site, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1, size = 8),
        strip.background = element_blank(),
        strip.text.x = element_blank()) +
  xlab("") +
  ylab("") +
  ylim(0,175)
zn_ms <-  annotate_figure(zn_ms,right = text_grob("Zn", color = "black", 
                                 face = "bold", size = 16, rot = 0))

# arrange 2 charts together
p <- ggarrange(cu_ms,zn_ms,nrow = 2)
annotate_figure(p,
                top = text_grob("Kenai River Mainstem Sites", 
                                color = "black", face = "bold", size = 18),
                left = text_grob("Concentration (ug/L)", color = "black", 
                                 face = "bold", size = 16, rot = 90),
                fig.lab = "— = CCC",
                fig.lab.pos = "bottom.right",
                fig.lab.face = "bold")


```

Metal observations were made but have missing CCC values for 5/22/2019 and
7/24/2019.  Find out why metal observations missing!

